Tracing the Meta-Level: PyPy's Tracing JIT Compiler

### 2009-04-02 talk/icooolps2009/paper.tex
Fix a couple of XXXs.

One of the often-cited drawbacks of dynamic languages is the performance
penalties they impose. Typically they are slower than statically typed languages
\cite{XXX}. Even though there has been a lot of research into improving the
performance of dynamic languages \cite{XXX}, those techniques are not as widely
used as one would expect. Many dynamic language implementations use completely
straightforward bytecode-interpreters without any advanced implementation
techniques like just-in-time compilation. There are a number of reasons for
this. Most of them boil down to the inherent complexities of using compilation.
Interpreters are simple to understand and to implement whereas writing a
just-in-time compiler is an error-prone task that is even made harder by the
dynamic features of a language.

writing an interpreter has many advantages... XXX

A recent approach to getting better performance for dynamic languages is that of
tracing JIT compilers. XXX [fijal] cite andreas gal paper?

@@

XXX contributions of this paper include:

@@

By writing VMs in a high-level language, we keep the implementation of the
language free of low-level details such as memory management strategy,
threading model or object layout.  These features are automatically added
during the translation process which consists in a series of steps, each step
transforming the representation of the program produced by the previous one
until we get the final executable.  As we will see later, this internal
low-level representation of the program is also used as an input for the
tracing JIT.

@@

At first those functions will be interpreted, but after a while, profiling shows
that the \texttt{while} loop in \texttt{strange\_sum} is executed often.  The
tracing JIT will then start trace the execution of that loop.  The trace would
look as follows:
\begin{verbatim}
loop_header(result0, n0)
i0 = int_mod(n0, Const(46))
i1 = int_eq(i0, Const(41))
guard_false(i1)
result1 = int_add(result0, n0)
n1 = int_sub(n0, Const(1))
i2 = int_ge(n1, Const(0))
guard_true(i2) [result1]
jump(result1, n1)
\end{verbatim}

XXX add a note about the SSA-ness of the trace

\fijal{Following paragraph is more confusing than helpful. The trace contains
all operations performed, including full inlining of functions called.
The loop will loop over the path in the interpreter, which corresponds to
path taken when tracing (which we assume is a common case).
Later, which is outside of the scope of this paper, less common cases
might be turned into assembler
code as well, producing bridges, if they're common enough. Note that this
loop is infinite, which means the only way to exit it is via guard failure}
This trace will then be turned into machine code. Note that the machine code
loop is by itself infinite and can only be left via a guard failure. Also note
\texttt{f} was inlined into the loop and how the common \texttt{else} case was
turned into machine code, while the other one is implemented via a guard
failure. The variables in square brackets after the guards are the state that
the interpreter will get when the guard fails.

@@

\fijal{I find following paragraph out of scope and completely confusing, we
should instead simply state that we unroll the loop, how we do that and
why we do that. Completely ignore aspect of an interpreter loop I suppose,
because everything previously keeps talking about can\_enter\_jit that closes
loop being available at jump back bytecodes}
A tracing JIT compiler finds the hot loops of the program it is compiling. In
our case, this program is the language interpreter. The hot loop of the language
interpreter is its bytecode dispatch loop. Usually that is is also the only hot
loop of the language interpreter \arigo{Uh?}.  Tracing one iteration of this loop means that
the recorded trace corresponds to execution of one opcode. This means that the
assumption that the tracing JIT makes -- that several iterations of a hot loop
take the same or similar code paths -- is just wrong in this case. It is very
unlikely that the same particular opcode is executed many times in a row.

@@

The first integration problem is how to \emph{not} integrate the tracing JIT at
all. It should be possible to choose when the interpreter is translated to C
whether the JIT should be built in or not. If the JIT is not enabled, all the
hints that are possibly in the interpreter source are just ignored by the
translation process. In this way, the result of the translation is identical to
as if no hints were present in the interpreter at all.

@@

Classical partial evaluation has failed to be useful for dynamic language for
much the same reasons why ahead-of-time compilers cannot compile them to
efficient code. If the partial evaluator knows only the program it simply does
not have enough information to produce good code. Therefore some work has been
done to do partial evaluation at runtime. One of the earliest works on runtime
specialisation is Tempo for C \cite{XXX}. However, it is essentially a normal
partial evaluator ``packaged as a library''; decisions about what can be
specialised and how are pre-determined. Another work in this direction is DyC
\cite{grant_dyc_2000}, another runtime specialiser for C. Both of these projects
have a similar problem as DynamoRIO.  Targeting the C language makes
higher-level specialisation difficult (e.g.\ \texttt{malloc} can not be
optimized).

>>>>

One of the often-cited drawbacks of dynamic languages is the performance
penalties they impose. Typically they are slower than statically typed languages
\cite{XXX}. Even though there has been a lot of research into improving the
performance of dynamic languages \cite{XXX}, those techniques are not as widely
used as one would expect. Many dynamic language implementations use completely
straightforward bytecode-interpreters without any advanced implementation
techniques like just-in-time compilation. There are a number of reasons for
this. Most of them boil down to the inherent complexities of using compilation.
Interpreters are simple to implement, understand, extend and port whereas writing a
just-in-time compiler is an error-prone task that is made even harder by the
dynamic features of a language.

A recent approach to getting better performance for dynamic languages is that of
tracing JIT compilers \cite{XXX}. Writing a tracing JIT compiler is relatively
simple, because it can be added to an existing interpreter for a language,
because the interpreter takes over some of the functionality of the compiler and
the machine code generation part can be simplified.

@@

The contributions of this paper are:
\begin{itemize}
\item Techniques for improving the generated code when applying a tracing JIT to
an interpreter
\item 
\end{itemize}

@@

By writing VMs in a high-level language, we keep the implementation of the
language free of low-level details such as memory management strategy,
threading model or object layout.  These features are automatically added
during the translation process. The process starts by performing control flow
graph construction and type inferences, then followed by a series of steps, each step
transforming the intermediate representation of the program produced by the
previous one until we get the final executable.  The first transformation step
makes details of the Python object model explicit in the intermediate
representation, later steps introducing garbage collection and other low-level
details. As we will see later, this internal representation of the program is
also used as an input for the tracing JIT.

@@

To trace this, a bytecode form of these functions needs to be introduced that
the tracer understands. The tracer interprets a bytecode that is an encoding of
the intermediate representation of PyPy's translation toolchain after type
inference has been performed and Python-specifics have been made explicit. At
first those functions will be interpreted, but after a while, profiling shows
that the \texttt{while} loop in \texttt{strange\_sum} is executed often.  The
tracing JIT will then start to trace the execution of that loop.  The trace would
look as follows:
\begin{verbatim}
loop_header(result0, n0)
i0 = int_mod(n0, Const(46))
i1 = int_eq(i0, Const(41))
guard_false(i1)
result1 = int_add(result0, n0)
n1 = int_sub(n0, Const(1))
i2 = int_ge(n1, Const(0))
guard_true(i2)
jump(result1, n1)
\end{verbatim}

The operations in this sequence are operations of the mentioned intermediate
representation (e.g. note that the generic modulo and equality operations in the
function above have been recognized to always work on integers and are thus
rendered as \texttt{int\_mod} and \texttt{int\_eq}). The trace contains all the
operations that were executed, is in SSA-form \cite{XXX} and ends with a jump
to its own beginning, forming an endless loop that can only be left via a guard
failure. The call to \texttt{f} was inlined into the trace. Of the condition in
\texttt{f} the much more common \texttt{else} case was traced. The other case is
implemented via a guard failure. This trace can then be turned into machine code
and executed.

@@

\fijal{I find following paragraph out of scope and completely confusing, we
should instead simply state that we unroll the loop, how we do that and
why we do that. Completely ignore aspect of an interpreter loop I suppose,
because everything previously keeps talking about can\_enter\_jit that closes
loop being available at jump back bytecodes}
A tracing JIT compiler finds the hot loops of the program it is compiling. In
our case, this program is the language interpreter. The most important hot loop
of the language interpreter is its bytecode dispatch loop (for many simple
interpreters it is also the only hot loops).  Tracing one iteration of this
loop means that
the recorded trace corresponds to execution of one opcode. This means that the
assumption that the tracing JIT makes -- that several iterations of a hot loop
take the same or similar code paths -- is just wrong in this case. It is very
unlikely that the same particular opcode is executed many times in a row.

@@

The first integration problem is how to \emph{not} integrate the tracing JIT at
all. It should be possible to choose when the language interpreter is translated to C
whether the JIT should be built in or not. If the JIT is not enabled, all the
hints that are possibly in the interpreter source are just ignored by the
translation process. In this way, the result of the translation is identical to
as if no hints were present in the interpreter at all.

@@

Classical partial evaluation has failed to be useful for dynamic language for
much the same reasons why ahead-of-time compilers cannot compile them to
efficient code. If the partial evaluator knows only the program it simply does
not have enough information to produce good code. Therefore some work has been
done to do partial evaluation at runtime. One of the earliest works on runtime
specialisation is Tempo for C \cite{XXX}. However, it is essentially a normal
partial evaluator ``packaged as a library''; decisions about what can be
specialised and how are pre-determined. Another work in this direction is DyC
\cite{grant_dyc_2000}, another runtime specializer for C. Both of these projects
have a similar problem as DynamoRIO.  Targeting the C language makes
higher-level specialisation difficult (e.g.\ \texttt{malloc} can not be
optimized).
